Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-16 _tex/index.tex
Built site for gh-pages

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(X\))
influence students' conceptual understanding of algebra (\(Y\))?'' To
investigate this, the study randomly assigns students to two groups,
each using different algebra teaching method \((X=\{1,2\})\). Students'
conceptual understanding of algebra is evaluated through pairwise
comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\),
indicating which student possesses a higher degree of understanding
among those compared. The research question is then framed as the
estimand, ``\emph{On average}, is there a difference in conceptual
understanding of algebra between the two groups of student?'' This
estimand is mathematically expressed as \(E[Y| X=1] - E[Y| X=2]\), where
\(E[\cdot]\) denotes the expected value. An example of this approach is
seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refer to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
estimator known for its effectiveness in modeling pairwise comparisons,
providing accurate estimates for the desired estimand when its
assumptions are met.

@@

Although experiments are widely regarded as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014},
researchers often encounter constraints that limit their ability to
conduct experimental studies. These constraints include ethical
concerns, such as the assignment of individuals to potentially harmful
interventions, and practical limitations, such as the infeasibility of,
for example, assigning individuals to genetic modifications or physical
impairments \citep{Neal_2020}. In these instances, causal inference
offers a valuable alternative for generating causal estimates.
\emph{Causal inference} is a framework designed to identify the causes
of phenomena and estimate their effects using data
\citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
modeling, which focuses primarily on summarizing data and inferring
associations, causal inference provides a rigorous mathematical
framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
Moreover, the framework is beneficial for both observational and
experimental studies, as it provides significant theoretical insights
that enhance their design \citep{McElreath_2020}.

Counterfactuals represent the highest level of cognitive abstraction in
the ladder of causation, followed by intervention and association
\citep{Pearl_et_al_2018}, and a detailed explanation of causes,
counterfactuals, and their applications is beyond the scope of this
document. However, it is important to note that causal inference
addresses the question, ``What would have happened in the world under
different circumstances?'' by transforming unobserved \emph{potential
outcomes}, known as counterfactuals, into associational estimates.
\emph{Counterfactuals} represent scenarios \emph{contrary to fact},
where alternative \emph{potential outcomes} resulting from a cause are
neither observed nor observable \citep{Neal_2020, Counterfactual_2024}.
The framework achieves this transformation while ensuring the absence of
confounding. Ultimately, the \emph{fundamental problem of causality} is
reduced to calculating differences between averages of observed outcomes
and the transformed counterfactuals, thereby identifying causal effects.

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
mathematical notation to characterize causal inference, but they do so
in different ways \citep{Neal_2020}. The potential outcomes approach
relies on counterfactual notation, whereas the structural approach
employs do-calculus \citep{Pearl_2009}. Despite these differences, both
notations can be expressed in terms of the other, and both approaches
provide methods for using experimental and observational data to
estimate causal effects \citep{Pearl_2010}.

However, the structural approach offers an additional key advantage over
the potential outcomes approach: it enables the graphical and formal
representation of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020} and and structural causal
models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. DAGs serve as
heuristics, effectively conveying the assumed causal structure of a
system \citep{Pearl_et_al_2016}. A heuristic is a strategy that
simplifies information to make decisions more quickly, efficiently, and
sometimes more accurately than complex methods \citep{Chow_2015}.
Consequently, DAGs do not represent detailed statistical models but
allow researchers to deduce which statistical models can provide valid
causal inferences, assuming the causal structure depicted in the DAGs
are accurate \citep{McElreath_2020}. The Identification-Estimation
flowchart in Figure~\ref{fig-IEflow} visually represents the process of
transitioning from estimands to estimates, and the application of the
sicentific model and data to identify and estimates causal effects.

>>>>

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(X\))
influence students' conceptual understanding of a topic (\(Y\))?'' To
investigate this, the study could randomly assign students to two
groups, each using a different teaching method \((X=\{1,2\})\).
Students' conceptual understanding of the topic could be evaluated
through pairwise comparisons, resulting in a dichotomous outcome
\((Y=\{0,1\})\), indicating which student among those compared has a
higher level of understanding. The research question could be then
framed as the estimand, ``\emph{On average}, is there a difference in
conceptual understanding of the topic between the two groups of
students?'' This estimand could be mathematically expressed as
\(E[Y| X=1] - E[Y| X=2]\), where \(E[\cdot]\) denotes the expected
value. An example of this approach is seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refer to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the BTL model \citep{Bradley_et_al_1952, Luce_1959} is an
estimator known for its effectiveness in modeling pairwise comparisons,
yielding accurate estimates for the target of inference discussed
earlier, as long as its assumptions are met.

@@

Experiments are widely regarded as the gold standard in evidence-based
science \citep{Hariton_et_al_2018, Hansson_2014}, however, researchers
often encounter constraints that limit their ability to conduct
experimental studies. These constraints include ethical concerns, such
as the assignment of individuals to potentially harmful interventions,
and practical limitations, such as the infeasibility of, for example,
assigning individuals to genetic modifications or physical impairments
\citep{Neal_2020}. In these instances, causal inference offers a
valuable alternative for generating causal estimates. \emph{Causal
inference} is a framework designed to identify the causes of phenomena
and estimate their effects using data
\citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical statistical
modeling, which focuses primarily on summarizing data and inferring
associations, causal inference provides a rigorous mathematical
framework for analyzing causes and counterfactuals \citep{Pearl_2009}.
Consequently, the framework offers significant theoretical insights that
enhance the design of observational and experimental studies
\citep{McElreath_2020}.

Although a detailed explanation of causes, and counterfactuals is beyond
the scope of this document, it suffices to say that causal inference
addresses the \emph{fundamental problem of causality} through the use of
counterfactuals \citep{Neal_2020}. The framework defines the
\emph{individual causal effect} (ICE) as the difference between an
individual's observed and unobserved potential outcomes. These
unobserved potential outcomes, termed \emph{counterfactuals}, represent
hypothetical scenarios that are \emph{contrary to fact}, where
alternative outcomes for an individual, resulting from a specific cause,
are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The framework then extends this
estimand to groups by introducing the \emph{average causal effect}
(ACE). The ACE reflects the difference between averages of observed
potential outcomes and counterfactuals across groups. Finally, similar
to an experiment, the framework identifies the ACE from associational
estimates by ensuring the absence of confounding.

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
mathematical notation to characterize the ACE, but they do so in
different ways \citep{Neal_2020}. The potential outcomes approach relies
on counterfactual notation, whereas the structural approach employs
do-calculus \citep{Pearl_2009}. Despite these differences, both
notations can be expressed in terms of the other, and both approaches
provide methods for using experimental and observational data to
estimate causal effects \citep{Pearl_2010}.

The structural approach, however, offers a notable advantage over the
potential outcomes approach by allowing the graphical and formal
representation of theories through directed acyclic graphs
\citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
DAGs function as heuristics, effectively conveying the presumed causal
structure of a system, referred to as a \emph{scientific model}
\citep{McElreath_2020}. They do not represent detailed statistical
models but allow researchers to deduce which statistical models can
provide valid causal inferences, assuming the causal structure depicted
in the DAGs are accurate \citep{McElreath_2020}. The
Identification-Estimation flowchart in Figure~\ref{fig-IEflow} visually
represents this process of transitioning from estimands to estimates, as
well as the application of the scientific model and data to identify and
estimate causal effects.
